%
\documentclass[]{article}
\usepackage{html}
\usepackage{amsmath,amssymb}
\usepackage{psfig,epsfig}
\usepackage{html}
\usepackage[T1]{fontenc}
\usepackage{rotating}
%
\newcommand{\command}{\textbf}      % for commands
\newcommand{\replaceable}{\textit}         % for replaceable names e.g. inputfiles
\newcommand{\flags}{\textmd}        % flags of files
\newcommand{\directory}{\textsl} % flags of files/code
\newcommand{\src}{\texttt}          % source code
\newcommand{\myemph}{\emph}         % emphasize
%
\begin{document}
%
\begin{titlepage}
%
\begin{latexonly}
\noindent {\bf High Altitude Observatory} \\
\noindent National Center for Atmospheric Research, Boulder, CO \\
\vspace{2in}
\end{latexonly}
%
\begin{center}
{\Large\bf TIEGCM V1.8 Model Description} \\
{\bf May 6, 2005}
\medskip
{\it Author: Astrid Maute} \\
{\it e-mail: maute@ucar.edu}
\end{center}
%
\end{titlepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagenumbering{Roman}
\setcounter{page}{1}
%
\include{Cover}
%
\tableofcontents
%
\include{notation}
%
%{List of Figures}
%
%{List of Tables}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\pagenumbering{arabic}
\newpage
%
\include{introduction}
%
\include{electro}
%
\include{input2electro}
%
\include{mapping}
%
\include{highlat_mod}
%
\include{fieldline}
%
% should be in the default code 
\include{pg_current}
%
\include{finitediff}
%
\include{solve}
%
\include{efield}
%
\section{Additional features}
%
\include{current4dB}
\subsection{No electrodynamo calculation}
%
\begin{thebibliography}{1}
\newcommand{\enquote}[1]{`#1'}
%
\bibitem[Peym93]{peym93}
{\sc Peymirat, C. and A. Richmond} [1993].
\newblock \enquote{Modeling the ion loss and the generation of region 2 field--aligned
currents via equivalent magnetospheric conductances.}
\newblock {\em J. Geop. Res.\/}, {\bf 98}, pp.~15,467--15,476.
%
\bibitem[Rich74]{rich74}
{\sc Richmond, A.} [1974].
\newblock \enquote{The computation of magnetic effects of field-aligned
  magnetospheric currents.}
\newblock {\em J. Atmos. Terr. Phys.\/}, {\bf 36}, pp.~245--252.
%
\bibitem[Rich95]{rich95}
{\sc Richmond, A.} [1995].
\newblock \enquote{Ionospheric electrodynamics using apex coordinates.}
\newblock {\em J. Geomag. Geoelectr.\/}, {\bf 47}, pp.~191--212.
%
\bibitem[Rich05]{rich05}
{\sc Richmond, A.} [2005].
\newblock \enquote{Modeling magnetic perturbation from TIEGCM.}
\newblock {\em in preparation\/}.
%
\end{thebibliography}
%
\end{document}
